---
title: "Income Tax and Budget Calculator"
author: "Nick Dylla"
output:
  html_document: default
---
#### Federal Income Tax for Single Filing Tax Bracket
```{r}
fed_tax_func <- function(salary,
         brackets = c(9525,38700,82500,157500,200000,500000,Inf),
         rates = c(0.10,0.12,0.22,0.24,0.32,0.35,0.37)) {        
    sum(diff(c(0, pmin(salary, brackets))) * rates)
}
# fed_tax_func(100000) #function works, double-checked online at ('https://www.taxact.com/tools/tax-bracket-calculator')

```

### Example of Flat-Rate Income Tax (Illinois)
#### Total Illinois Income Tax for Single Filing Tax Bracket
```{r}
il_budget <- function(salary = data){
   std_deduct = 12000 # 2019 standard deduction is 12000
   fed_tax = fed_tax_func(salary-std_deduct) # federal tax with standard deduction included
   no_deduc_fed_tax = fed_tax_func(salary) # federal tax without standard deduction included
   state_tax = 0.0495*salary # flat rate
   soc_sec = 0.0620*salary # flat rate
   medicare = 0.0145*salary # flat rate
   tot_tax = fed_tax+state_tax+soc_sec+medicare
   no_deduc_tot_tax = no_deduc_fed_tax+state_tax+soc_sec+medicare
   post_tax = salary - no_deduc_tot_tax
   post_deduct_tax = salary - tot_tax
   return(data.frame('Salary'=salary,'Federal Tax'=no_deduc_fed_tax,'State Tax'=state_tax, 
                     'Social Security'=soc_sec, 'Medicare'=medicare,'Total Tax'=tot_tax,
                     'Income After Tax w/o Deduction'=post_tax,
                     'Income After Tax w/ Deduction'=post_deduct_tax))
}
il <- il_budget(100000) # Test run
il # Works out
```

### Example of Bracketed State Income Tax (Missouri)
#### Missouri Income Tax for Single Filing Tax Bracket
```{r}
mo_tax_func <- function(salary,
         brackets = c(1008,2016,3024,4032,5040,6048,7056,8064,9072,Inf),
         rates = c(0.015,0.02,0.025,0.03,0.035,0.04,0.045,0.05,0.055,0.060)) {        
    sum(diff(c(0, pmin(salary, brackets))) * rates)
}
```

#### Total Missouri Income Tax for Single Filing Tax Bracket
```{r}
mo_budget <- function(salary = data){
   std_deduct = 12000 # 2019 standard deduction is 12000
   fed_tax = fed_tax_func(salary-std_deduct) # federal tax with standard deduction included
   no_deduc_fed_tax = fed_tax_func(salary) # federal tax without standard deduction included
   state_tax = mo_tax_func(salary)
   soc_sec = 0.0620*salary
   medicare = 0.0290*salary
   tot_tax = fed_tax+state_tax+soc_sec+medicare
   no_deduc_tot_tax = no_deduc_fed_tax+state_tax+soc_sec+medicare
   post_tax = salary - no_deduc_tot_tax
   post_deduct_tax = salary - tot_tax
   return(data.frame('Salary'=salary,'Federal Tax'=no_deduc_fed_tax,'State Tax'=state_tax, 
                     'Social Security'=soc_sec, 'Medicare'=medicare,'Total Tax'=tot_tax,
                     'Income After Tax w/o Deduction'=post_tax,
                     'Income After Tax w/ Deduction'=post_deduct_tax))
}
mo <- mo_budget(100000) # Test run
mo # Works out
```

#### Compare Incomes Between Illinois and Missouri for Single Filing Tax Bracket
```{r}
rbind(il,mo)
```

#### User-defined Living Expenses
##### This function is meant for inputting personal/ variable costs
```{r}
custom_living_expenses_func <- function(salary=salary, rent=0,rental_insur = 0,
                                        utilities = 0,internet = 0, online_streaming = 0, 
                                        cell_phone = 0, groceries = 0,
                                        eat_out = 0, entertainment = 0, clothing = 0,
                                        miscellaneous = 0, life_insur = 0, dental_insur = 0, 
                                        vision_insur = 0, health_insur = 0, Rx_insur = 0,
                                        accident_insur = 0, car_insur = 0, gasoline = 0,
                                        loan = 0, one_time_pet_cost = 0, pet_food = 0,
                                        annual_pet_exam = 0, pet_license = 0, 
                                        gym_membership = 0){
  total_cost = rent+rental_insur+utilities+internet+online_streaming+cell_phone+groceries+eat_out+
                      entertainment+clothing+miscellaneous+life_insur+dental_insur+vision_insur+
                      health_insur+Rx_insur+accident_insur+car_insur+gasoline+loan+
                      one_time_pet_cost+pet_food+annual_pet_exam+pet_license+gym_membership
  savings = salary-total_cost
  return(data.frame(salary,total_cost,savings))
} # This function is not great for inputting personal/ variable costs due to the set values
```

#### Total Cost of Living (2 Salaries)
##### User-defined 'custom_living_expenses_func'
```{r}
# Salary 1 = 75000
# Salary 2 = 25000
sal_1 <- il_budget(75000)
sal_2 <- il_budget(25000)
total_sal <- sal_1$Income.After.Tax.w..Deduction+sal_2$Income.After.Tax.w..Deduction # columns containing the salary after taxes with the standard deduction included
custom_living_expenses_func(salary=total_sal,rent = 18000) # This will return the combined salaries, total living expense costs, and combined savings if you only had to pay rent 
```

